A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Additional Methods

### 2019-08-20 writeUp/supplementary_materials_methods.tex
Updated POA

\section{Detinova parity analysis}\label{sec:detinova}
\cite{detinova1962age} provided an alternative approach to estimate the lifespan of individual specimens, based on sampling the parous rate in a population (the proportion of adult females that have laid eggs) using dissection to determine parity of each specimen. In what follows, we follow the derivation given by \cite{davidson1954estimation}, to determine an expression for the proportion of nulliparous mosquitoes in terms of the gonotrophic cycle length $G$, here assumed the same for each cycle subsequent to an adult emerging. If a constant daily survival rate $p$ is assumed, the proportion of emerging adults surviving to become parous is $p^G$. If the rate of recruitment into the adult classes is constant over time, the proportion of nulliparous mosquitoes is,
%
\begin{equation}\label{eq:detinova}
\begin{aligned}
\mathbb{n} &= \frac{p^G}{p^G + p^{2G} + p^{3G} + \cdots}\\
&= p^G / \frac{p^G}{1-p^G}\\
&= 1-p^G.
\end{aligned}
\end{equation}
%
The daily survival rate can hence be calculated by inverting eq. (\ref{eq:detinova}): $p = \sqrt[G]{\mathcal{P}}$, where $\mathcal{P}=1-\mathbb{n}$ is the proportion parous sampled from the population.

@@

We estimated lifespan at the species and complex / subgroup level (hereafter denoted as \textit{s.l.}) using a Bayesian hierarchical framework. We denote this grouping variable by $s$ in what follows. The number of parous individuals $X_i^{\{s\}}$ in sample $i$ (consisting of individuals belonging to group $s$) was modelled by a binomial sampling distribution,
%
\begin{equation}
X_i^{\{s\}} \sim \mathcal{B}(N, \mathcal{P}_i^{\{s\}}),
\end{equation}
%
where $N$ is the sample size and $\mathcal{P}_i^{\{s\}}$ is the proportion parous in sample $i$ for group $s$. A hierarchical prior structure was used for the parous proportion in each sample,
%
\begin{equation}
\mathcal{P}_i^{\{s\}}\sim \text{beta}\left(\phi^{\{s\}} \kappa^{\{s\}}, (1-\phi^{\{s\}})\kappa^{\{s\}}\right),
\end{equation}
%
where $0\leq\phi^{\{s\}}\leq 1$ is the mean parous proportion and $\kappa^{\{s\}}>1$ is a concentration parameter (both of which are group-specific).

The priors we used in this model were: $\phi\sim \text{beta}(20, 14)$ and $\kappa\sim \text{Pareto}(1, 1.5)$. These were chosen because this yielded a prior predictive distribution with a median value close to 10 (see Fig. \ref{fig:detinova_priors}).

@@

>>>>

\section{Detinova parity analysis}\label{sec:detinova}
\cite{detinova1962age} provided an alternative approach to estimate the lifespan of individual specimens, based on sampling the parous rate in a population (the proportion of adult females that have laid eggs) using dissection to determine parity of each specimen. In what follows, we follow the derivation given by \cite{davidson1954estimation}, to determine an expression for the proportion of nulliparous mosquitoes in terms of the gonotrophic cycle length $G$, here assumed the same for each cycle subsequent to an adult emerging. If a constant daily survival rate $p$ is assumed, the proportion of emerging adults surviving to become parous is $p^G$. If the rate of recruitment into the adult classes is constant over time, the proportion of nulliparous mosquitoes is,
%
\begin{equation}\label{eq:detinova}
\begin{aligned}
\mathbb{n} &= \frac{p^G}{p^G + p^{2G} + p^{3G} + \cdots}\\
&= p^G / \frac{p^G}{1-p^G}\\
&= 1-p^G.
\end{aligned}
\end{equation}
%
The daily survival rate can hence be calculated by inverting eq. (\ref{eq:detinova}): $p = \sqrt[G]{\mathcal{P}}$, where $\mathcal{P}=1-\mathbb{n}$ is the proportion parous sampled from the population. The distribution of mosquitoes dying each day follows a geometric distribution, with probability of dying on a given day equal to $1-p$. The mean lifespan is, hence, given by $1/(1-p)$.

@@

We estimated lifespan at the species and complex / subgroup level (hereafter denoted as \textit{s.l.}) using a Bayesian hierarchical framework. We denote this grouping variable by $s$ in what follows. The number of parous individuals $X_i^{\{s\}}$ in sample $i$ (consisting of individuals belonging to group $s$) was modelled by a binomial sampling distribution,
%
\begin{equation}
X_i^{\{s\}} \sim \mathcal{B}(N_i, \mathcal{P}_i^{\{s\}}),
\end{equation}
%
where $N_i$ is the sample size and $\mathcal{P}_i^{\{s\}}$ is the proportion parous in sample $i$ for group $s$. A hierarchical prior structure was used for the parous proportion in each sample,
%
\begin{equation}\label{eq:detinova_prior}
\mathcal{P}_i^{\{s\}}\sim \text{beta}\left(\phi^{\{s\}} \kappa^{\{s\}}, (1-\phi^{\{s\}})\kappa^{\{s\}}\right),
\end{equation}
%
where $0\leq\phi^{\{s\}}\leq 1$ is the mean parous proportion and $\kappa^{\{s\}}>1$ is a concentration parameter (both of which are group-specific).

The priors we used in this model were: $\phi\sim \text{beta}(20, 14)$ and $\kappa\sim \text{Pareto}(1, 1.5)$. These were chosen because this yielded a prior predictive distribution with a median lifespan close to 10 (see Fig. \ref{fig:detinova_priors}).

@@

To investigate the effect of measures of temperature, we extracted daily data from the ERA-interim re-analysis by ECMWF \citep{dee2011era} through their website using the Python API. This data included:

\begin{itemize}
	\item Two-meter surface temperature of air at 2m above the surface of land, sea or in-land waters.
	\item Daily minimum two-meter surface temperature.
	\item Daily maximum two-meter surface temperature.
\end{itemize}

For this analysis, we included only those studies where the start month listed in the database coincided with the end date - meaning the study took place within a given month - resulting in $n=742$ data points. The temperature data was extracted for the GPS position of each study ($\pm 0.1 \text{ degrees}$) for each day of the month over which mosquito sampling took place, with a grid size of 0.05 degrees. The Python script to extract this information is given below.

\begin{minted}{python}
from ecmwfapi import ECMWFDataServer
import time
def extract_meanminmax_temperature(a_area, a_date_start,
		a_date_end):
  server = ECMWFDataServer()
  date_range = str(a_date_start) + "/to/" + str(a_date_end)
  date_file = str(a_date_start) + "_to_" + str(a_date_end)
  area=(str(a_area["north"]) + '/' + str(a_area["west"]) +
       '/' + str(a_area["south"]) + '/' + str(a_area["east"])) 
  area_file = (str(a_area["north"]) + '_' + str(a_area["west"])
  + '_' + str(a_area["south"]) + '_' + str(a_area["east"]))
  a_filename = ("../data/raw/weather/temperature_" +
  	str(area_file) + "_" + date_file + ".nc")
  server.retrieve({
    "class": "ei",
    "dataset": "interim",
    "date": str(date_range),
    "expver": "1",
    "grid": "0.05/0.05",
    "levtype": "sfc",
    "area" : str(area),
    "param": "167.128/201.128/202.128",
    "step": "3",
    "stream": "oper",
    "time": "00:00:00",
    "type": "fc",
    "target":  str(a_filename),
    "format" : "netcdf",
  })
\end{minted}

To determine the impact of temperature on mosquito lifespan, we estimated a generalised linear model using a logit link, with hierarchical complex-level intercepts and either no temperature term, and linear temperature terms with or without quadratic temperature terms. In each case, these temperature terms were allowed to vary at the complex level (non-hierarchically). This meant our likelihood had the form,
%
\begin{equation}
X_i^{\{s\}} \sim \mathcal{B}(N_i, \text{inv-logit}(\alpha_i^{\{s\}})),
\end{equation}
%
where,
%
\begin{equation}
\alpha_i^{\{s\}} = \text{logit} (\mathcal{P}_i^{\{s\}}) + \delta_1^{\{s\}} T_i + \delta_2^{\{s\}} T_i^2.
\end{equation}
%
Here $T_i$ is a measure of temperature at study location $i$ (measures described below); $\mathcal{P}_i$ was set a prior according to eq. \eqref{eq:detinova_prior} and $\delta_1^{\{s\}}\sim \mathcal{N}(0, 1)$ and $\delta_2^{\{s\}}\sim \mathcal{N}(0, 1)$.

The temperature measures ($T_i$), we investigated were summary statistics calculated across the daily data for all days within the month each study took place and across all grid points falling within the area centred on the GPS location of each study (described above),

\begin{itemize}
	\item Average two meter temperature (average of mean daily data for each grid point),
	\item Minimum two meter temperature (average of minimum daily data for each grid point),
	\item Maximum two meter temperature (average of maximum daily data for each grid point),
	\item Standard deviation in two meter temperature (average of standard deviation in temperature for each grid point),
	\item Daily range in two meter temperature (average of maximum minus minimum daily temperature for each grid point).
\end{itemize}
